---
title: "TCGA PRAD"
author: "Peter Chernek and Levi Waldron"
date: "December 23, 2015"
output: pdf_document
---

# Load / install packages

```{r, message=FALSE}
##BiocInstaller::biocLite("lwaldron/LeviRmisc", dependencies=TRUE)
library(LeviRmisc)
library(MultiAssayExperiment)
library(DESeq2)
library(Biobase)
library(edgeR)
```


# Download TCGA PRAD data

```{r, cache=FALSE}
if(!file.exists("prad_rnaseq_eset.rds") | !file.exists("prad_mirnaseq_eset.rds")){
  download.file("http://s3.amazonaws.com/multiassayexperiments/pradMAEO.rds", destfile="pradMAEO.rds")
  maeo <- readRDS("pradMAEO.rds")
}
```

```{r}
pdat <- read.csv("https://raw.githubusercontent.com/lwaldron/tcga_prad/master/prad_clinical.csv", row.names=1)
rownames(pdat) <- toupper(rownames(pdat))
```

# Create `ExpressionSet` objects for RNA-seq and miRNA-seq

RNA-seq:
```{r rnaseq, cache=FALSE}
if(!file.exists("prad_rnaseq_eset.rds") | !file.exists("prad_mirnaseq_eset.rds")){
  eset.rnaseq <- experiments(maeo[, , "RNASeq2GeneNorm"])[[1]]
  colnames(eset.rnaseq) <- sub("[AB]$", "", colnames(eset.rnaseq))
  eset.rnaseq <- eset.rnaseq[, match(rownames(pdat), colnames(eset.rnaseq))]
  all.equal(rownames(pdat), colnames(eset.rnaseq))
  pData(eset.rnaseq) <- pdat
  eset.rnaseq <- eset.rnaseq[, grep("01$", colnames(eset.rnaseq))]
  eset.rnaseq <- eset.rnaseq[, !is.na(eset.rnaseq$racevar)]
  eset.rnaseq <- eset.rnaseq[, eset.rnaseq$racevar %in% c("black or african american", "white")]
  eset.rnaseq$racevar <- factor(eset.rnaseq$racevar, levels=c("white", "black or african american"))
  write.csv(exprs(eset.rnaseq), file="prad_rnaseq.csv")
  write.csv(pData(eset.rnaseq), file="prad_clinical.csv")
  saveRDS(eset.rnaseq, file="prad_rnaseq_eset.rds")
}
```

miRNA-seq:
```{r mirnaseq, cache=FALSE}
if(!file.exists("prad_rnaseq_eset.rds") | !file.exists("prad_mirnaseq_eset.rds")){
  eset.mirnaseq <- experiments(maeo[, , "miRNASeqGene"])[[1]]
  colnames(eset.mirnaseq) <- sub("[AB]$", "", colnames(eset.mirnaseq))
  pdat2 <- pdat[rownames(pdat) %in% colnames(eset.mirnaseq), ]
  eset.mirnaseq <- eset.mirnaseq[, match(rownames(pdat2), colnames(eset.mirnaseq))]
  all.equal(rownames(pdat2), colnames(eset.mirnaseq))
  pData(eset.mirnaseq) <- pdat2
  eset.mirnaseq <- eset.mirnaseq[, grep("01$", colnames(eset.mirnaseq))]
  eset.mirnaseq <- eset.mirnaseq[, grep("01$", colnames(eset.mirnaseq))]
  eset.mirnaseq <- eset.mirnaseq[, !is.na(eset.mirnaseq$racevar)]
  eset.mirnaseq <- eset.mirnaseq[, eset.mirnaseq$racevar %in% c("black or african american", "white")]
  eset.mirnaseq$racevar <- factor(eset.mirnaseq$racevar, levels=c("white", "black or african american"))
  write.csv(exprs(eset.mirnaseq), file="prad_mirnaseq.csv")
  write.csv(pData(eset.mirnaseq), file="prad_clinical.csv")
  saveRDS(eset.mirnaseq, file="prad_mirnaseq_eset.rds")
}
```


To load the esets again:

```{r, cache=TRUE}
eset.rnaseq <- readRDS("prad_rnaseq_eset.rds")
eset.mirnaseq <- readRDS("prad_mirnaseq_eset.rds")
```

# Create files for Broad GSEA tool

```{r, cache=TRUE}
LeviRmisc::makeCLS(eset.rnaseq$racevar, fname="race.cls")
output <- cbind(rownames(eset.rnaseq), rep(NA, nrow(eset.rnaseq)), log2(exprs(eset.rnaseq) + 1))
colnames(output)[1:2] = c("NAME", "DESCRIPTION")
write.table(output, file="race.gct", sep="\t", quote=FALSE)
```

# Some random analyses that could be deleted

## Make a histogram of PSA (KLK3) expression

```{r}
hist(exprs(eset.rnaseq["KLK3", ]))
hist(log(exprs(eset.rnaseq["KLK3", ])))
```

## See what clinical data are available by default:

```{r, eval=FALSE}
summary(pData(eset.rnaseq))
```

## Look for association between KLK3 expression and clinical PSA

```{r}
psadat <- data.frame(psa=as.numeric(as.character(eset.rnaseq$patient.stage_event.psa.psa_value)),
                     klk3=t(exprs(eset.rnaseq["KLK3", ])))
psadat.complete <- psadat[complete.cases(psadat), ]
plot(KLK3 ~ psa, data=psadat.complete, xlab="clinical PSA", ylab="KLK3 tumor expression", log="xy")
fit <- lowess(x=psadat.complete$psa, y=psadat.complete$KLK3)
lines(fit, col="red", lw=3)
```

## Is there an association between PSA in the urine and KLK3 in the tumor?

```{r}
cor.test(x=psadat$KLK3,  y=psadat$psa, method="spearman")
```


## Clinical psa as a function of ethnicity

```{r}
table(eset.rnaseq$racevar)
boxplot(psadat$psa ~ eset.rnaseq$race, ylab="PSA")
```


# miRNA differential expression

```{r, cache=TRUE}
dge <- DGEList(counts=exprs(eset.mirnaseq))
## Require at least 5 counts in at least 5% of samples:
dge$counts <- dge$counts[(apply(dge$counts, 1, function(x) sum(x > 5)) > ncol(dge$counts) * 0.05), ]
design = model.matrix(~ batch_number + gleason_score + racevar, 
                      data=pData(eset.mirnaseq))
v <- voom(dge, design, plot=TRUE)
fit <- lmFit(v, design)
fit <- eBayes(fit)
tt <- topTable(fit, coef="racevarblack or african american")
write.csv(tt, file="voom_mirnaseq.csv")
```

## Heatmap

Correcting for batch and Gleason score.

```{r heatmapmirna}
library(pheatmap)
sig.mirnaseq <- rownames(tt)[tt$adj.P.Val < 0.05]
log2.norm.counts <- log2(exprs(eset.mirnaseq) + 1)[sig.mirnaseq, ]
df <- pData(eset.mirnaseq)[, c("gleason_score", "racevar"), drop=FALSE]
raceorder <- order(df$racevar, colSums(log2.norm.counts))
df <- df[raceorder, , drop=FALSE]
log2.norm.counts <- log2.norm.counts[, raceorder]
stopifnot(identical(rownames(df), colnames(log2.norm.counts)))
pheatmap(log2.norm.counts, cluster_rows=TRUE, show_colnames=FALSE, 
         cluster_cols=TRUE,
         clustering_distance_rows="correlation", 
         clustering_distance_cols="correlation", 
         scale="none", annotation_col=df["racevar"])
```

Boxplot of the top differentially expressed miRNA
```{r mirnaboxplot}
boxplot(log2.norm.counts[sig.mirnaseq[1], ] ~ df$racevar,
        xlab="Race", ylab="log2 expression", main=sig.mirnaseq[1])
stripchart(log2.norm.counts[sig.mirnaseq[1], ] ~ df$racevar, vertical=TRUE,
           method="jitter",
            pch = 21, col = "maroon", bg = "bisque", 
            add = TRUE)
```

# RNA-seq differential expression

Correcting for batch and Gleason score.

```{r, cache=TRUE}
dge <- DGEList(counts=exprs(eset.rnaseq))
## Require at least 5 counts in at least 5% of samples:
dge$counts <- dge$counts[(apply(dge$counts, 1, function(x) sum(x > 5)) > ncol(dge$counts) * 0.05), ]
design = model.matrix(~ batch_number + gleason_score + racevar, 
                      data=pData(eset.rnaseq))
v <- voom(dge, design, plot=TRUE)
fit <- lmFit(v, design)
fit <- eBayes(fit)
tt <- topTable(fit, coef="racevarblack or african american")
write.csv(tt, file="voom_rnaseq.csv")
```

## Heatmap

```{r heatmaprna}
library(pheatmap)
sig.rnaseq <- rownames(tt)[tt$adj.P.Val < 0.05]
log2.norm.counts <- log2(exprs(eset.rnaseq) + 1)[sig.rnaseq, ]
df <- pData(eset.rnaseq)[, c("gleason_score", "racevar"), drop=FALSE]
raceorder <- order(df$racevar, colSums(log2.norm.counts))
df <- df[raceorder, , drop=FALSE]
log2.norm.counts <- log2.norm.counts[, raceorder]
stopifnot(identical(rownames(df), colnames(log2.norm.counts)))
pheatmap(log2.norm.counts, cluster_rows=TRUE, show_colnames=FALSE, 
         cluster_cols=TRUE,
         clustering_distance_rows="correlation", 
         clustering_distance_cols="correlation", 
         scale="none", annotation_col=df["racevar"])
```

Boxplot of the top differentially expressed rna
```{r rnaboxplot}
boxplot(log2.norm.counts[sig.rnaseq[1], ] ~ df$racevar,
        xlab="Race", ylab="log2 expression", main=sig.rnaseq[1])
stripchart(log2.norm.counts[sig.rnaseq[1], ] ~ df$racevar, vertical=TRUE,
           method="jitter",
            pch = 21, col = "maroon", bg = "bisque", 
            add = TRUE)
```



# Spearman correlation between miR-1304 and some mRNA

```{r, warning=FALSE}
int = intersect(sampleNames(eset.mirnaseq), sampleNames(eset.rnaseq))
goi = featureNames(eset.rnaseq)
goi = c("FBXO45", "ACBD3", "PFKFB2", "USP47", "MKX", "CAPRIN2", "AKR1B1", "KIAA1324")
goi = c("KIAA1324", "AKR1B1", "CAPRIN2", "MKX")
eset.mirnaseq2 = eset.mirnaseq["hsa-mir-1304", int]
eset.rnaseq2 = eset.rnaseq[goi, int]
tests=apply(exprs(eset.rnaseq2), 1, function(x){
    res=cor.test(x, exprs(eset.mirnaseq2)[1, ], method="spearman")
    output = c(res$estimate, res$p.value)
    names(output) = c("rho", "p-value")
    return(output)
})
tests=t(tests)
tests = tests[order(tests[, 2]), ]
tests
```
